Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing

### 2019-10-21 Dissonance_Santiago_v1.Rmd
Worked on Analysis and Results, included Cochran-Mantel-Haenszel Chi-Squared Test for 3-way tables

---
title: Do Drivers Dream of Walking? An Investigation of Dissonance in the Primary Mode of Travel from the Perspective of Subjective Wellbeing
author:
  - name: Author 1
    email: a1@university.com
    affiliation: Some University
    footnote: Corresponding Author
  - name: Author 2
    email: a2@school.com
    affiliation: Some School
address:
  - code: Some University
    address: Department, Street, City, State, Zip
  - code: Some School
    address: Department, Street, City, State, Zip
abstract: |
  Transportation in most of the world has been dominated for decades by a fascination with the automobile. Increasingly, however, it is recognized that to achieve a variety of economic, environmental, and public health policy goals it is important to attract and retain active travelers and users of public transportation. A challenge with the way transportation policy is developed is that it tends to focus on utilitarian considerations, which may miss other potential aspects of transportation that users value. In particular, subjective wellbeing (SWB) has been proposed as a way to enhance our understanding of the preferences and choices of travellers, as well as a way to evaluate the benefits of transportation beyond utilitarian considerations. The objective of this paper is to analyze the modes that people commonly use, and to what extent they are aligned (or not) with a variety of affective values associated with SWB. In other words, we are interested in the potential for dissonance with respect to the primary mode of travel, from the perspective of affective and utilitarian values. The study is based on data collected from a sample of travellers in the city of Santiago, in Chile. Participants in the study were asked about their usual mode of travel, and then were asked to name the mode or modes that they associated to affective values freedom, enjoyment, happiness, poverty, luxury and status. The results indicate that users of public transportation experience the most dissonance in terms of affective values, and active travellers the least. For those travellers who experience dissonance, active travel is the mode most commonly associated with freedom, enjoyment, and happiness, public transportation is most commonly associated with poverty, and the automobil is most commonly associated with luxury and status.

@@

library(rticles)
library(readxl)
library(tidyverse)
library(rgdal)
library(sf)
library(ggmosaic)
library(ggthemes)
library(kableExtra)
```

@@

# To run this notebook, the file `Dissonance_Santiago.RData` is needed. This file contains the following object:
# 
# master: a dataframe with information collected from 451 respondents. The variables in the dataframe are:
# 
# - NUMERO: Respondent ID
# - USED: Main mode of transportation used by the respondent
# - FREEDOM: Mode(s) that the respondent associates with feelings of freedom
# - ENJOYMENT: Mode(s) that the respondent associates with feelings of enjoyment
# - HAPPINESS: Mode(s) that the respondent associates with feelings of happiness
# - POVERTY: Mode(s) that the respondent associates with feelings of poverty
# - LUXURY: Mode(s) that the respondent associates with feelings of luxury
# - STATUS: Mode(s) that the respondent associates with feelings of status
# - SAFETY: Mode(s) that the respondent associates with feelings of safety
# - TIME: Mode(s) that the respondent associates with feelings of WASTED TIME
# - EFFICIENCY: Mode(s) that the respondent associates with feelings of efficiency
# - AGE: Age of the respondent
# - EDUCATION: Level of education of the respondent
# - INCOME: Level of income of respondent
# - TRAVEL_TIME: Trip duration of regular commuting trip 
# - WEIGHT_*: A weight for the responses by ATTRIBUTE. Some respondents selected more than one mode that they associated with feelings of {FREEDOM, ENJOYMENT, HAPPINESS, POVERTY, LUXURY, STATUS, SAFETY, TIME, and EFFICIENCY}. To avoid giving greater weight to respondents who indicated several modes of transportation, a weight was calculated that is the inverse of the number of modes indicated. These weights can be used to weight the frequencies prior to any analysis.
# 
# The modes of transportation (USED, FREEDOM, ENJOYMENT, HAPPINESS, POVERTY, LUXURY, STATUS, SAFETY, TIME, EFFICIENCY) are coded as: 
# "A": Auto
# "B": Taxi
# "C": Colectivo
# "D": Moto
# "E": Metro
# "F": Bus
# "G": Bicycle
# "H": Walk
#
# AGE is coded as:
#
# "A": Less18
# "B": 18to24
# "C": 25to34
# "D": 35to54
# "E": 55to64
# "F": 65more
# 
# EDUCATION is coded as:
# 
# "A": ELEMENTARY
# "B": SECUNDARY
# "C": PROFESS_TECH
# "D": COLLEGE
# "E": POSTGRAD
#
# INCOME is coded as (in pesos?):
# 
# "A": Less423
# "B": 423to639
# "C": 639to977
# "D": 977to1550
# "E": 1550to2380
# "F": 2380more
# 
# TRAVEL_TIME is coded as:
#
# "A": 0-20 min
# "B": 20-40 min
# "C": 40-60 min
# "D": 1h and more

@@

# The master dataframe is processed for analysis. We plan to conduct two distinct types of analysis:
#
# 1. Analysis of dissonace using the characteristics of respondents (i.e., AGE, EDUCATION, INCOME, (Mode) USED, and TRAVEL_TIME) by ATTRIBUTE (i.e., FREEDOM et al.). In this case, dissonance is defined as follows:
#
# >> If the mode used is identical to the mode that the respondent associates with ATTRIBUTE (i.e. FREEDOM et al.), we consider that there is no dissonance; otherwise, we consider that there is dissonance. For example, if the respondent's main mode of transportation is Auto, and the respondent associates Auto with feelings of FREEDOM, then there is no dissonance. On the other hand, if the respondent's main mode of transportation is Auto, but associates Walking with feelings of FREEDOM, there is dissonance.
# 
# Given this, we will calculate how frequently there is dissonance at different levels of AGE, EDUCATION, INCOME, MODE, TRAVEL_TIME. Tests of independence can be implemented.
#
# 2. Analysis of dissonance using the mode USED and the mode respondents associated with ATTRIBUTE, then stratified by characteristics of respondents (i.e., AGE et al.) in this case, we see how frequently each mode was associated with ATTRIBUTE, given the main mode of transportation of the respondents.
#
# Here the dataframes are prepared to support these two types of analysis

#################################################################################
# Prepare a dataframe for the analysis of dissonance by attribute (Analysis 1.) #
#################################################################################

@@

Overall_dissonance <- Overall_dissonance %>%
  mutate(USED = case_when(USED == "A" ~ "Auto",
                                  USED == "B" | USED == "C" | USED == "D" ~ "Other",
                                  USED == "E" | USED == "F" ~ "Public",
                                  USED == "G" | USED == "H" ~ "Active"),
         AGE = case_when(AGE == "A" | AGE == "B" | AGE == "C" ~ "< 35",
                         AGE == "D" ~ "35 - 54",
                         AGE == "E" | AGE == "F" ~ "> 54"),
         EDUCATION = case_when(EDUCATION == "A" | EDUCATION == "B" ~ "K-12",
                               EDUCATION == "C" | EDUCATION == "D" ~ "Tech/Univ",
                               EDUCATION == "E" ~ "Graduate"),
         INCOME = case_when(INCOME == "A" | INCOME == "B" ~ "Low",
                            INCOME == "C" | INCOME == "D" ~ "Middle",
                            INCOME == "E" | INCOME == "F" ~ "High"),
         TRAVEL_TIME = case_when(TRAVEL_TIME == "A" ~ "< 20 min",
                                 TRAVEL_TIME == "B" ~ "20 - 40 min",
                                 TRAVEL_TIME == "C" ~ "40 - 60 min",
                                 TRAVEL_TIME == "D" ~ "> 60 min")) %>%
  mutate(USED = factor(USED, levels = c("Auto", "Active", "Public", "Other"), labels = c("Auto", "Active", "Public", "Other")),
         FREEDOM_DISSONANCE = factor(FREEDOM_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         ENJOYMENT_DISSONANCE = factor(ENJOYMENT_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         HAPPINESS_DISSONANCE = factor(HAPPINESS_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         POVERTY_DISSONANCE = factor(POVERTY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         LUXURY_DISSONANCE = factor(LUXURY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         STATUS_DISSONANCE = factor(STATUS_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         SAFETY_DISSONANCE = factor(SAFETY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         TIME_DISSONANCE = factor(TIME_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         EFFICIENCY_DISSONANCE = factor(EFFICIENCY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         AGE = factor(AGE, levels = c("< 35", "35 - 54", "> 54"), labels = c("< 35", "35 - 54", "> 54")),
         EDUCATION = factor(EDUCATION, levels = c("K-12", "Tech/Univ", "Graduate"), labels = c("K-12", "Tech/Univ", "Graduate")),
         INCOME = factor(INCOME, levels = c("Low", "Middle", "High"), labels = c("Low", "Middle", "High")),
         TRAVEL_TIME = factor(TRAVEL_TIME, levels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min"), labels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min")))

#############################################################################
# Prepare a dataframe for the analysis of dissonance by modes (Analysis 2.) #
#############################################################################

@@

Modes_by_Attribute <- Modes_by_Attribute %>%
  mutate(USED = case_when(USED == "A" ~ "Auto",
                                  USED == "B" | USED == "C" | USED == "D" ~ "Other",
                                  USED == "E" | USED == "F" ~ "Public",
                                  USED == "G" | USED == "H" ~ "Active"),
         IDEAL = case_when(IDEAL == "A" ~ "Auto",
                                  IDEAL == "B" | IDEAL == "C" | IDEAL == "D" ~ "Other",
                                  IDEAL == "E" | IDEAL == "F" ~ "Public",
                                  IDEAL == "G" | IDEAL == "H" ~ "Active"),
         AGE = case_when(AGE == "A" | AGE == "B" | AGE == "C" ~ "< 35",
                         AGE == "D" ~ "35 - 54",
                         AGE == "E" | AGE == "F" ~ "> 54"),
         EDUCATION = case_when(EDUCATION == "A" | EDUCATION == "B" ~ "K-12",
                               EDUCATION == "C" | EDUCATION == "D" ~ "Tech/Univ",
                               EDUCATION == "E" ~ "Graduate"),
         INCOME = case_when(INCOME == "A" | INCOME == "B" ~ "Low",
                            INCOME == "C" | INCOME == "D" ~ "Middle",
                            INCOME == "E" | INCOME == "F" ~ "High"),
         TRAVEL_TIME = case_when(TRAVEL_TIME == "A" ~ "< 20 min",
                                 TRAVEL_TIME == "B" ~ "20 - 40 min",
                                 TRAVEL_TIME == "C" ~ "40 - 60 min",
                                 TRAVEL_TIME == "D" ~ "> 60 min"))

@@

Modes_by_Attribute <- Modes_by_Attribute %>%
  mutate(USED = factor(USED, levels = c("Auto", "Active", "Public", "Other"), labels = c("Auto", "Active", "Public", "Other")),
         IDEAL = factor(IDEAL, levels = c("Auto", "Active", "Public", "Other"), labels = c("Auto", "Active", "Public", "Other")),
         AGE = factor(AGE, levels = c("< 35", "35 - 54", "> 54"), labels = c("< 35", "35 - 54", "> 54")),
         EDUCATION = factor(EDUCATION, levels = c("K-12", "Tech/Univ", "Graduate"), labels = c("K-12", "Tech/Univ", "Graduate")),
         INCOME = factor(INCOME, levels = c("Low", "Middle", "High"), labels = c("Low", "Middle", "High")),
         TRAVEL_TIME = factor(TRAVEL_TIME, levels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min"), labels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min")))
```

@@

Introduction
==========================

Transportation planning for decades has focused on providing mobility for the private automobile. This is a model of development that was initially introduced in North America as a solution to urbanizations problems, and that was eventually copied elsewhere (Brown et al., 2009; Angotti, 1996) . Despite the initial promise of automotive technology, it is now evident that mobility centred on the private automobile has given rise to a litany of maladies that are in urgent need of correction. This includes environmental concerns (i.e., climate change; Chapman, 2007) as well as other social (Boschmann and Kwan, 2008), health (Khreis et al., 2016), and equity issues (Bocarejo and Oviedo, 2012). 
As the real impacts of our societal dependence on the private autombobile have become increasingly evident, the transportation agenda has in recent years aimed to shift focus to the reduction of car-use and towards the creation of mobility polycultures that offer a broader menu of transportation alternatives than just the private automobile (Miller, 2011) . In order to successfully achieve this goal, it is essential not only to provide the services and facilities that support public transportation and active travel, but also to attract new users to these modes of transportation (Ettema et al. 2011). Within this context, it has been argued that sustainable transportation policies require all participants in the transportation system to challenge what Gossling and Cohen (2014) term “transportation taboos”: deep-seated ideas concerning the contribution to emissions by individuals, the inequality of market-based approaches, and the social and psychological functions of transportation. With respect to the latter, moving beyond a purely utilitarian focus, it is important to understand how users perceive affective and instrumental elements on their transportation experience, since this can help to incorporate adequate improvements in alternative transportation, including transit and active modes, to make them more attractive to users (Anable and Gatersleben, 2005; Domarchi et al., 2008; Gatersleben and Uzzell, 2007; Steg 2005 ).

A useful lens to investigate the social and psychological functions of transportation is by means of the concept of subjective wellbeing (SWB. See, inter alia, De Vos et al., 2013 and Chatterjee et al., 2019) . SWB is defined by the OECD as “[g]ood mental states, including all of the various evaluations, positive and negative, that people make of their lives, and the affective reactions of people to their experiences” (OECD, 2013, p. 10).  Accordingly, the aim of this paper is to analyse some of the factors that can potentially affect the levels of SWB in the wide spectrum of public, private and active transport modes in the context of Santiago de Chile. Previous research suggests that public transport users experience a negative gap in desired and actual travel time, whereas active travellers don’t seem to mind somewhat longer trips than what strictly needed (e.g., Páez and Whalen, 2010). In the context of a Latin American country, historically the poor travel experience of public transit users and active travellers reflects the social inequalities of social groups that have no alternative but to move in these types of modes while living far from the work centres and main activities. The paper aims to reflect on the possibilities of increasing the SWB while travelling, and how this could increase the attractiveness of public transport and active travel in the socio-demographic groups studied. This can help not only to reduce the use of private transport but also decrease the inequality gap between these users.

The paper contributes to the literature in three areas. First, it contributes to an emerging literature on the topic in the context of the Global South (Bejarano et al., 2017 ; Shao and Liang, 2019 ; Van et al., 2014 ; Zorrilla et al., 2019);  to the best of our knowledge, the case of Chile has not yet been reported. Second, although there is literature on the enjoyment of commute and other emotional factors, from a hedonic and even eudaimonic perspectives, the analysis has yet to be applied more fully in terms of distributional issues – i.e. which groups are mostly disadvantaged. Finally, the analysis shows the attitudes of people towards their primary mode and their perception towards ‘ideal modes’ – implying their preferences, even in situations when their ideal mode is not part of their actual choice set. The attitudes of travellers towards transport modes are critical factors to be considered by policy-makers in case they want to promote and increase the use of public transport or active modes (e.g., Bejarano et al., 2019). The paper explores these positive and negative associations within the transport options used in Santiago.

Background
============

The transportation community has for the most part already agreed to the need of looking at mobility and transport issues from the lens of wellbeing, to broaden the commonly used utilitarian perspective to evaluate projects. Users’ travel satisfaction (Bergstad, 2001) has been the primary way to direct this challenge towards everyday mobility, resulting on research about satisfaction with different transport modes. For example, numerous studies report that car users have a higher level of satisfaction compared to other transport modes (Redmond and Mokhtarian, 2001 ; Bergstad et al. 2011; Eriksson et al. 2013; Whalen et al., 2013). In a similar way, there are multiple reports that active travel also tends to yield high levels of satisfaction (Smith, 2017; St-Louis et al. 2014; Paez and Whalen, 2010; Whalen et al., 2013), In contrast, public transport users tend to assess their experience more negatively (De Vos et al. 2016; Abenoza et al. 2017; Ettema et al. 2012; Páez and Whalen, 2010). Multi-modal trips also influence satisfaction levels; for instance, when an individual chooses a particular mode of transportation, she will report a higher level of satisfaction with that chosen mode – perhaps as a form of post hoc validation (Susilo and Cats, 2014).

While the use of travel satisfaction has been mainly used in the context of daily trips – typically been linked to cost-benefit and utilitarian measurements –, the evaluation of Subjective Wellbeing (SWB) over time has risen as an alternative measure. In the field of travel behaviour, Ettema et al. (2010; p. 725) have defined SWB as the degree to which an individual positively evaluates the overall quality of their lives, where the general life satisfaction encompasses a more extended temporality – which implies assuming a tendency to be more stable over time. This concept has prompted a growing literature that complements and applies SWB in a broader range of satisfaction scales and situations. The definition of other factors such as travel choice mode, attitudes and external elements of the built environment has been studied for a broader understanding of the changes produced in the SWB. As these factors do not necessarily apply to the general life satisfaction on the long term, the studies have aimed to determine both the direct and indirect effects on the perception of users (see, e.g. Ye and Titheridge 2017). Other concepts have also emerged as the Satisfaction with Travel Scale (STS), a measurement devised by Ettema et al. (2011) as well as different scales based on people’s travel perceptions. De Vos et al. (2015) for instance, explore in detail the underlying dimensions of the affective domain of STS in which SWB is based (for more on STS see also Friman et al. 2013).

>>>>

---
title: Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing
author:
  - name: Author 1
    email: a1@university.com
    affiliation: Some University
    footnote: Corresponding Author
  - name: Author 2
    email: a2@school.com
    affiliation: Some School
address:
  - code: Some University
    address: Department, Street, City, State, Zip
  - code: Some School
    address: Department, Street, City, State, Zip
abstract: |
  Transportation in most of the world has been dominated for decades by a fascination with the automobile. Nonetheless, there is an increasing recognition that to achieve a variety of economic, environmental, and public health policy goals it is important to attract and retain active travelers and users of public transportation. A challenge with the way transportation policy is developed, however, is that it tends to focus on utilitarian considerations, which may miss other potential aspects of transportation that users value. In particular, subjective wellbeing (SWB) has been proposed as a way to enhance our understanding of the preferences and choices of travellers, as well as a way to evaluate the benefits of transportation beyond utilitarian considerations. The objective of this paper is to analyze the modes that people commonly use, and to what extent they are aligned (or not) with a variety of affective values associated with SWB. In other words, we are interested in the potential for dissonance with respect to the primary mode of travel, from the perspective of affective values. The study is based on data collected from a sample of travellers in the city of Santiago, in Chile. Participants in the study were asked about their usual mode of travel, and then were asked to name the mode or modes that they associate with the affective values of freedom, enjoyment, happiness, poverty, luxury and status. The results indicate that users of public transportation experience the most dissonance in terms of affective values, and active travellers the least. For those travellers who experience dissonance, active travel is the mode most commonly associated with freedom, enjoyment, and happiness, public transportation is most commonly associated with poverty, and the automobil is most commonly associated with luxury and status.

@@

library(rticles)
library(readxl)
library(tidyverse)
library(rgdal)
library(sf)
library(ggmosaic)
library(ggthemes)
library(kableExtra)
library(gridExtra)
```

@@

# To run this notebook, the file `Dissonance_Santiago.RData` is needed. This file contains the following object:
# 
# master: a dataframe with information collected from 451 respondents. The variables in the dataframe are:
# 
# - NUMERO: Respondent ID
# - USED: Main mode of transportation used by the respondent
# - FREEDOM: Mode(s) that the respondent associates with feelings of freedom
# - ENJOYMENT: Mode(s) that the respondent associates with feelings of enjoyment
# - HAPPINESS: Mode(s) that the respondent associates with feelings of happiness
# - POVERTY: Mode(s) that the respondent associates with feelings of poverty
# - LUXURY: Mode(s) that the respondent associates with feelings of luxury
# - STATUS: Mode(s) that the respondent associates with feelings of status
# - SAFETY: Mode(s) that the respondent associates with feelings of safety
# - TIME: Mode(s) that the respondent associates with feelings of WASTED TIME
# - EFFICIENCY: Mode(s) that the respondent associates with feelings of efficiency
# - AGE: Age of the respondent
# - EDUCATION: Level of education of the respondent
# - INCOME: Level of income of respondent
# - TRAVEL_TIME: Trip duration of regular commuting trip 
# - WEIGHT_*: A weight for the responses by ATTRIBUTE. Some respondents selected more than one mode that they associated with feelings of {FREEDOM, ENJOYMENT, HAPPINESS, POVERTY, LUXURY, STATUS, SAFETY, TIME, and EFFICIENCY}. To avoid giving greater weight to respondents who indicated several modes of transportation, a weight was calculated that is the inverse of the number of modes indicated. These weights can be used to weight the frequencies prior to any analysis.
# 
# The modes of transportation (USED, FREEDOM, ENJOYMENT, HAPPINESS, POVERTY, LUXURY, STATUS, SAFETY, TIME, EFFICIENCY) are coded as: 
# "A": Car
# "B": Taxi
# "C": Colectivo
# "D": Moto
# "E": Metro
# "F": Bus
# "G": Bicycle
# "H": Walk
#
# AGE is coded as:
#
# "A": Less18
# "B": 18to24
# "C": 25to34
# "D": 35to54
# "E": 55to64
# "F": 65more
# 
# EDUCATION is coded as:
# 
# "A": ELEMENTARY
# "B": SECUNDARY
# "C": PROFESS_TECH
# "D": COLLEGE
# "E": POSTGRAD
#
# INCOME is coded as (in pesos?):
# 
# "A": Less423
# "B": 423to639
# "C": 639to977
# "D": 977to1550
# "E": 1550to2380
# "F": 2380more
# 
# TRAVEL_TIME is coded as:
#
# "A": 0-20 min
# "B": 20-40 min
# "C": 40-60 min
# "D": 1h and more

@@

# The master dataframe is processed for analysis. We plan to conduct two distinct types of analysis:
#
# 1. Who experiences dissonace?
#
# Analysis of dissonace using the characteristics of respondents (i.e., AGE, EDUCATION, INCOME, (Mode) USED, and TRAVEL_TIME) by ATTRIBUTE (i.e., FREEDOM et al.). In this case, dissonance is defined as follows:
#
# >> If the mode used is identical to the mode that the respondent associates with ATTRIBUTE (i.e. FREEDOM et al.), we consider that there is no dissonance; otherwise, we consider that there is dissonance. For example, if the respondent's main mode of transportation is Car, and the respondent associates Car with feelings of FREEDOM, then there is no dissonance. On the other hand, if the respondent's main mode of transportation is Car, but associates Walking with feelings of FREEDOM, there is dissonance.
# 
# Given this, we will calculate how frequently there is dissonance at different levels of AGE, EDUCATION, INCOME, MODE, TRAVEL_TIME. Tests of independence can be implemented.
#
# 2. For those respondents who experience dissonance, what mode(s) are better aligned with various affective values
#
# Analysis of dissonance using the mode USED and the mode respondents associated with ATTRIBUTE, then stratified by characteristics of respondents (i.e., AGE et al.) in this case, we see how frequently each mode was associated with ATTRIBUTE, given the main mode of transportation of the respondents.
#
# Here the dataframes are prepared to support these two types of analysis

#####################################################################################
# Prepare a dataframe for Analysis 1: Who experiences dissonance by affective value #
#####################################################################################

@@

Overall_dissonance <- Overall_dissonance %>%
  mutate(USED = case_when(USED == "A" ~ "Car",
                                  USED == "B" | USED == "C" | USED == "D" ~ "Other",
                                  USED == "E" | USED == "F" ~ "Public",
                                  USED == "G" | USED == "H" ~ "Active"),
         AGE = case_when(AGE == "A" | AGE == "B" | AGE == "C" ~ "< 35",
                         AGE == "D" ~ "35 - 54",
                         AGE == "E" | AGE == "F" ~ "> 54"),
         EDUCATION = case_when(EDUCATION == "A" | EDUCATION == "B" ~ "K-12",
                               EDUCATION == "C" | EDUCATION == "D" ~ "Tech/Univ",
                               EDUCATION == "E" ~ "Graduate"),
         INCOME = case_when(INCOME == "A" | INCOME == "B" ~ "Low",
                            INCOME == "C" | INCOME == "D" ~ "Middle",
                            INCOME == "E" | INCOME == "F" ~ "High"),
         TRAVEL_TIME = case_when(TRAVEL_TIME == "A" ~ "< 20 min",
                                 TRAVEL_TIME == "B" ~ "20 - 40 min",
                                 TRAVEL_TIME == "C" ~ "40 - 60 min",
                                 TRAVEL_TIME == "D" ~ "> 60 min")) %>%
  mutate(USED = factor(USED, levels = c("Car", "Active", "Public", "Other"), labels = c("Car", "Active", "Public", "Other")),
         FREEDOM_DISSONANCE = factor(FREEDOM_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         ENJOYMENT_DISSONANCE = factor(ENJOYMENT_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         HAPPINESS_DISSONANCE = factor(HAPPINESS_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         POVERTY_DISSONANCE = factor(POVERTY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         LUXURY_DISSONANCE = factor(LUXURY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         STATUS_DISSONANCE = factor(STATUS_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         SAFETY_DISSONANCE = factor(SAFETY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         TIME_DISSONANCE = factor(TIME_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         EFFICIENCY_DISSONANCE = factor(EFFICIENCY_DISSONANCE, levels = c(1, 2), labels = c("No Dissonance", "Dissonance")),
         AGE = factor(AGE, levels = c("< 35", "35 - 54", "> 54"), labels = c("< 35", "35 - 54", "> 54")),
         EDUCATION = factor(EDUCATION, levels = c("K-12", "Tech/Univ", "Graduate"), labels = c("K-12", "Tech/Univ", "Graduate")),
         INCOME = factor(INCOME, levels = c("Low", "Middle", "High"), labels = c("Low", "Middle", "High")),
         TRAVEL_TIME = factor(TRAVEL_TIME, levels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min"), labels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min")))

###############################################################################################
# Prepare a dataframe for Analysis 2: What mode(s) align better with various affective values #
###############################################################################################

@@

Modes_by_Attribute <- Modes_by_Attribute %>%
  mutate(USED = case_when(USED == "A" ~ "Car",
                                  USED == "B" | USED == "C" | USED == "D" ~ "Other",
                                  USED == "E" | USED == "F" ~ "Public",
                                  USED == "G" | USED == "H" ~ "Active"),
         IDEAL = case_when(IDEAL == "A" ~ "Car",
                                  IDEAL == "B" | IDEAL == "C" | IDEAL == "D" ~ "Other",
                                  IDEAL == "E" | IDEAL == "F" ~ "Public",
                                  IDEAL == "G" | IDEAL == "H" ~ "Active"),
         AGE = case_when(AGE == "A" | AGE == "B" | AGE == "C" ~ "< 35",
                         AGE == "D" ~ "35 - 54",
                         AGE == "E" | AGE == "F" ~ "> 54"),
         EDUCATION = case_when(EDUCATION == "A" | EDUCATION == "B" ~ "K-12",
                               EDUCATION == "C" | EDUCATION == "D" ~ "Tech/Univ",
                               EDUCATION == "E" ~ "Graduate"),
         INCOME = case_when(INCOME == "A" | INCOME == "B" ~ "Low",
                            INCOME == "C" | INCOME == "D" ~ "Middle",
                            INCOME == "E" | INCOME == "F" ~ "High"),
         TRAVEL_TIME = case_when(TRAVEL_TIME == "A" ~ "< 20 min",
                                 TRAVEL_TIME == "B" ~ "20 - 40 min",
                                 TRAVEL_TIME == "C" ~ "40 - 60 min",
                                 TRAVEL_TIME == "D" ~ "> 60 min"))

@@

Modes_by_Attribute <- Modes_by_Attribute %>%
  mutate(USED = factor(USED, levels = c("Car", "Active", "Public", "Other"), labels = c("Car", "Active", "Public", "Other")),
         IDEAL = factor(IDEAL, levels = c("Car", "Active", "Public", "Other"), labels = c("Car", "Active", "Public", "Other")),
         AGE = factor(AGE, levels = c("< 35", "35 - 54", "> 54"), labels = c("< 35", "35 - 54", "> 54")),
         EDUCATION = factor(EDUCATION, levels = c("K-12", "Tech/Univ", "Graduate"), labels = c("K-12", "Tech/Univ", "Graduate")),
         INCOME = factor(INCOME, levels = c("Low", "Middle", "High"), labels = c("Low", "Middle", "High")),
         TRAVEL_TIME = factor(TRAVEL_TIME, levels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min"), labels = c("< 20 min", "20 - 40 min", "40 - 60 min", "> 60 min")))
```

@@

1. Introduction
==========================

Transportation planning for decades has focused on providing mobility for the private automobile. This is a model of development that was initially introduced in North America as a solution to problems created by rapid urbanization, and that was eventually copied elsewhere [@Brown2009planning;@Angotti1996latin]. Despite the initial promise of automotive technology, it is now evident that mobility centred on the private automobile has given rise to a litany of maladies that are in urgent need of correction. This includes environmental concerns [i.e., climate change; @Chapman2007transport] as well as numerous other social [@Boschman2008toward; @Lucas2012transport; @Lucas2019evolution], health [@Khreis2016health; @Milne2012public], and equity issues [@Bocarejo2012transport; @Martens2012justice; @Pereira2017distributive].

As the real impacts of our societal dependence on the private autombobile have become increasingly evident, the transportation agenda has aimed to shift focus to the reduction of car use and towards the creation of mobility polycultures that offer a broader menu of transportation alternatives than primarily (or even just) the private automobile [@Miller2011collaborative]. In order to successfully achieve this goal, it is essential not only to provide the services and facilities that support public transportation and active travel, but also to attract new users to these modes of transportation [@Ettema2011satisfaction]. Within this context, it has been argued that sustainable transportation policies require all participants in the transportation system to challenge what Gossling and Cohen [-@Gossling2014sustainable] termed _transportation taboos_: deep-seated ideas concerning the contribution to emissions by individuals, the inequality of market-based approaches, and the social and psychological functions of transportation. With respect to the latter, it is important to move beyond a purely utilitarian focus if we are to understand how the affective value of the transportation experience can be leveraged to improve alternative transportation - particularly transit and active modes [@Anable2005work; @Domarchi2008effect; @Gatersleben2007affective; @Steg2005car].

A useful lens to investigate the social and psychological functions of transportation is by means of the concept of subjective wellbeing [SWB; see, inter alia, @Devos2013travel; and @Chatterjee2019commuting]. SWB is defined by the OECD as “[g]ood mental states, including all of the various evaluations, positive and negative, that people make of their lives, and the affective reactions of people to their experiences” (**OECD, 2013, p. 10**). As memorably put by Steg [-@Steg2005car]: is the use of car a must, or is it a lust? A close alignment, or consonance, between affective values and the mode of transportation used to travel can result in greater subjective wellbeing and increase the probability of choosing that mode  (i.e., by satisfying the _lust_); on the other hand, dissonance, that is, the lack of concordance between the mode used and affective values would be detrimental to subjective wellbeing, especially of those for whom the use of the mode is a _must_ [@Devos2019satisfaction; @Mokhtarian2018travel]. With these considerations in mind, one aim of the present research is to investigate _who_ experiences dissonance with respect to their primary mode of travel and affective values, namely freedom, enjoyment, happiness, feelings of poverty, luxury, and status. Furthermore, it is possible, even if use of a particular mode is a _must_, that travellers may still _lust_ for something else - in other words, the grass may in fact look greener from the window of the car. For this reason, we also aim to investigate which mode or modes are most commonly associated with affective values, with a focus on those travellers who experience dissonance in their primary mode.

Research is based on data collected from a sample of travellers in the city of Santiago in Chile. Survey respondents were asked about their primary mode of travel, and also about the mode or modes that they associate with the affective values mentioned above. The paper contributes to the literature in three ways, as follows:

1. The research reported here contributes to an emerging literature on the topic of transportation and affective values in the context of the Global South [@Alayyash2019commute; @Bejarano2017user; @Shao2019analysis; @Van2014effect; @Zorrilla2019exploring]; to the best of our knowledge, the case of Chile has not yet been reported.

2. Although there is an extensive literature on the enjoyment of commute and other affective values [see for instance @Paez2010enjoyment; @Redmond2001positive; @Whalen2013mode; @Ye2017satisfaction], from a hedonic and even eudaimonic perspectives the analysis has yet to be applied more fully in terms of distributional issues – i.e. which groups more commonly experience dissonance [see @Devos2018people].

3. The analysis shows the attitudes of people towards their primary mode and their perception towards ‘ideal modes’ – implying their preferences, even in situations when their ideal mode is not part of their actual choice set. More concretely, the results indicate that users of public transportation experience the most dissonance in terms of affective values, and active travellers the least. For those travellers who experience dissonance, active travel is the mode most commonly associated with freedom, enjoyment, and happiness, public transportation is most commonly associated with poverty, and the automobil is most commonly associated with luxury and status. We also find that there are some substantial variations in dissonance by age, education, income, and typical commute time. The attitudes of travellers towards transport modes are critical factors to be considered by policy-makers in case they want to promote and increase the use of public transport or active modes [@Bornioli2019affective; @Devos2017travel; @Devos2019satisfying; @Garling2019role; @Redman2013quality].

2. Background
============

A consensus has emerged in the transportation community regarding the need to complement the traditional utilitarian perspective of transportation by looking at mobility and transport issues from the lens of their affective functions. The affective value of transportation in turn is important due to its potential to improve or detract from SWB. One of the primary ways to explore this has been the satisfaction that travelers feel towards their every day mobility experience [e.g., @Bergstad2011subjective]. As a consequence, there is a wealth of research on satisfaction with the use of different modes of transportation. For example, numerous studies report that car users often have a higher level of satisfaction compared to other transport modes [@Bergstad2011affective; @Eriksson2013perceived; @Redmond2001positive; @Whalen2013mode; but see @Handy2019commute]. In a similar way, there are multiple reports that active travel also tends to yield high levels of satisfaction [@Gatersleben2007affective; @Smith2017commute; @Stlouis2014happy; @Paez2010enjoyment; @Whalen2013mode; @Handy2019commute], In contrast, public transport users tend to assess their experience more negatively [@Abenoza2017travel; @Devos2016travel; @Gatersleben2007affective; @Handy2019commute; @Paez2010enjoyment]. Multi-modal trips also influence satisfaction levels; for instance, when an individual chooses a particular mode of transportation, she will report a higher level of satisfaction with that chosen mode – perhaps as a form of _post hoc_ validation [@Susilo2014exploring].
